Data Provenance for Concurrent Program Debugging and Analysis in Production Systems

### 2013-11-12 paper.tex
committing a draft with more discussion of the main table in eval

We illustrate the debugging benefits of last writer slices by using our system
to debug several real-world programs studied previously in the debugging
literature.  Table~\ref{tab:bugs} provides an overview of the bugs we studied
describing the program, bug type, and bug report identifier where applicable.
We then provide a summary of last writer slicing's debugging benefits.  To do
so, we mirror the evaluation strategy of bad value origin
tracking~\cite{badapples}, as it is the only other low-overhead provenance
tracking mechanism we know of.  We evaluate our system using their three
evaluation criteria - (1) effectiveness, (2) triviality, and (3) usefulness.
Our technique is effective if the last writer slice reports a code point that
wrote a value that subsequently caused a failure.  A failure is trivially
debuggable if its cause is made obvious by examining the core dump alone, {\em
i.e.}, a last writer slice is unnecessary.  A last writer slice is useful if it
is effective and the thread and code point in the slice actually help
understand why the failure occurred or how to fix it.   To draw contrast to bad
value origin tracking, we indicate whether their analysis could help debug each
bug, based on the description of their technique.

\begin{table}
\scriptsize
\centering
\begin{tabular}{l|ll|l|ccc|c}
{\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    & Yes   & Yes \\
httrack& 3.43.9  &  N/A     & Order  &Yes  &No    & Yes   & No  \\%Used w/o origin - no help
jdk1.4 &  1.70   &          & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
\end{tabular}
\caption{{\bf Debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
\normalsize
\end{table}

\paragraph{Apache}
%http://issues.apache.org/bugzilla/show_bug.cgi?id=21287
%concurrency-bugs
We conducted a case study using Apache httpd-2.0.48, which has a bug in its
script cache. The cache uses reference counting to determine when to deallocate
cached objects.  When a thread sees a reference count hit zero, it frees the
object.  The bug is that the check that the count is zero and the deallocation
are not atomic but it should be.  Multiple threads may see the zero reference
count and free the object more than once, leading to a crash.

We debugged this issue by examining the core dump from a crashing execution.
The core dump showed that the memory allocator aborted due to one thread
freeing already freed memory at {\tt mod\_mem\_cache.c}, line 287.  We examined
the last writer table entry for the object being freed at that point.  That
entry revealed that a different thread last wrote that object at the same point
in the code.  The information in the last writer table illustrates why the
double free bug is happening -- two threads end up in the object deallocation
routine for the same object, which should not happen.

@@

\paragraph{PBZip2}
We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
a use-after-delete bug that leads to an assertion failure.  A worker thread
tries to acquire a mutex after it has been deleted by the main thread during
shutdown, causing the assertion to fail.

Last writer slices provide the key piece of information to debug this failure.
When the program fails, the core dump shows that the mutex is invalid because
it has been deleted.  The mutex's last writer slice shows that the last write
to the mutex before the worker tried to acquire it was when the main thread
deleted it.  The last writer slice reveals the connection between the the two
pieces of code involved in the bug: the code that acquires the mutex and the
code that deletes the mutex.  Note that this connection is not obvious when the
program crashes because the main thread may have executed arbitrarily far from
where it deleted the mutex before the worker thread crashes.

@@

We used last writer slices to debug an atomicity violation in MySQL-4.0.12 that
is a critical security vulnerability.  Figure~\ref{fig:mysqllws} shows the
buggy code.  Thread 1 rotates the database's log file, marking it closed,
opening a new log file, then marking it open again.  Log rotation should be
atomic, but the lack of synchronization allows Thread 2 to run its insert
during log rotation.  The insert code checks if the log is marked closed. If
the log is marked closed, the insert is not logged, manifesting the failure.
We added an {\tt abort} statement to the non-logging case because this behavior
is a security vulnerability that the bug report describes as ``critical''.
Adding this {\tt abort} is reasonable, as the bug report describes the
condition in the {\tt else} block as being an error.

@@

We triggered the bug by running an insert query during a log rotation.  The bug
manifested, stopping execution at {\bf C}.  As described in the bug report, the
query went unlogged and examining the core dump revealed the log was marked
closed at point {\bf B} ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
However, the core dump did not reveal {\em why} the log was marked closed
there.  The last writer slice at {\bf C} for the {\tt log\_type} variable shows
that it was written last by Thread 1 at {\bf A}.  Without the last writer
slice, the relationship between the log rotation code and the insert code is
not apparent. The last writer slice reveals that the log rotation code
interacts with the insert code incorrectly via the {\tt log\_type} variable.

\paragraph{JDK-StringBuffer}

The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:debugging} describes the benefit provided by
last writer slices in debugging an ordering violation in the JDK-1.4 stringbuffer library.

@@

{\bf TODO: Compare with these analyses implemented using binary instrumentation}

@@

{\bf TODO: Run without escape analysis?}

>>>>

We illustrate the debugging benefits of last writer slices by using our system
to debug six real-world programs studied previously in the debugging
literature.  Table~\ref{tab:bugs} provides an overview of this evaluation and
Section~\ref{sec:eval:debugging:cases} looks in detail at several cases to 
better illustrate how last writer slices helps with debugging.

In Table~\ref{tab:bugs}, we show the bugs we studied describing the program,
bug type, and bug report identifier where applicable.  For each bug, we
followed the procedure for installation and bug triggering described by the
author of~\cite{concurrencybugs}.  We study a mixture of concurrency bug types
in this work, including both atomicity and ordering violations.

The table also summarizes our evaluation of last writer slicing's debugging
benefits.  We mirror the evaluation strategy of bad value origin
tracking~\cite{badapples}, as it is the only other low-overhead provenance
tracking mechanism we know of.  We evaluate our system using their three
evaluation criteria - (1) {\bf effectiveness}, (2) {\bf triviality}, and (3)
{\bf usefulness}.  Our technique is {\bf effective} if the last writer slice
reports a code point that wrote a value that subsequently caused a failure.  A
failure is {\bf trivial} to debug if its cause is made obvious by examining the
core dump alone, {\em i.e.}, a last writer slice is unnecessary.  A last writer
slice is {\bf useful} if it is effective and the thread and code point in the
slice help understand why the failure occurred or how to fix it.   We directly
compare to bad value origin tracking.    The {\bf O.T.} column in
Table~\ref{tab:bugs} shows  whether their analysis could help debug each bug,
based on the description of their technique.

\begin{table}
\scriptsize
\centering
\begin{tabular}{l|ll|l|ccc|c}
{\bf App} & {\bf Ver. \#} & {\bf Bug \#} &{\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} & {\bf O.T.?}\\ \hline
mysql  &  4.0.12 &  791     & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
pbzip2 &  0.9.4  &  N/A     & Order  &Yes  &No    & Yes   & Yes \\
httrack& 3.43.9  &  N/A     & Order  &Yes  &No    & Yes   & No  \\%Used w/o origin - no help
jdk1.4 &  1.70   &          & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
apache1&  2.0.48 &  21285   & Atom.  &Yes  &No    & Yes   & No  \\%All values usable
apache2&  2.2.9  &  45605   & Atom.  &Yes  &No    & Maybe & ??  \\
\end{tabular}
\caption{\label{tab:bugs}{\bf Concurrency debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.  }
\normalsize
\end{table}

The results in Table~\ref{tab:bugs} show that for both atomicity and ordering
violation bugs, last writer slices are effective and useful.  In all cases, we
found that the last writer slice revealed the provenance of the data involved
in the operation that caused the failure, so last writer slicing was effective.
We found that none of the failures we looked at were trivially debuggable.
This finding is consistent with the difficulty of debugging concurrency bugs --
the core dump does not reveal the interaction between threads that led to a
failure, only the result of that interaction that leads to the failure.

Last writer slices were obviously useful in five of six cases.  In these cases,
the root cause of the bug becomes clear when last writer slices show the
inter-thread data flow led to a variable having a value that caused a failure.
Using that information, we expect it would be straightforward in these cases to
add synchronization to fix these bugs.  Note that without last writer slices,
understanding the responsible inter-thread data flow and fixing the bug would
be difficult.  For the remaining case, {\tt apache2}, last writer slices would
probably help with debugging, but it is less obvious because this bug is a
complex, multi-variable bug.  We provide more detail on {\tt apache2} in Section~\ref{sec:eval:debugging:cases}.

Our evaluation shows that bad value origin tracking~\cite{badapples} is not
likely to be helpful in five out of six cases where last writer slices is
useful.  In one case, a use-after-delete bug in {\tt pbzip2}, bad value origin
tracking helps.  For this bug, when one thread deletes a variable, it makes
that variable unusable and bad value origin tracking would track that value.
When the failure occurs, the programmer would then know why the value was
deleted.  In the other five cases we looked at, the values involved in causing
the failure are usable and would not be tracked by bad value origin tracking.

\subsubsection{Case Studies}
\label{sec:eval:debugging:cases}

We now provide detailed cases studies showing the debugging role of last writer
slices in several of the bugs we examined.  Space constraints prevent us from
providing details for all cases.

\paragraph{JDK-StringBuffer}
The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:debugging} describes the benefit provided by
last writer slices in debugging an ordering violation in the JDK-1.4.

%\paragraph{Apache}
%%http://issues.apache.org/bugzilla/show_bug.cgi?id=21287
%%concurrency-bugs
%We conducted a case study using Apache httpd-2.0.48, which has a bug in its
%script cache. The cache uses reference counting to determine when to deallocate
%cached objects.  When a thread sees a reference count hit zero, it frees the
%object.  The bug is that the check that the count is zero and the deallocation
%are not atomic but it should be.  Multiple threads may see the zero reference
%count and free the object more than once, leading to a crash. 
%
%We debugged this issue by examining the core dump from a crashing execution.
%The core dump showed that the memory allocator aborted due to one thread
%freeing already freed memory at {\tt mod\_mem\_cache.c}, line 287.  We examined
%the last writer table entry for the object being freed at that point.  That
%entry revealed that a different thread last wrote that object at the same point
%in the code.  The information in the last writer table illustrates why the
%double free bug is happening -- two threads end up in the object deallocation
%routine for the same object, which should not happen.

@@

%
%\paragraph{PBZip2}
%We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
%a use-after-delete bug that leads to an assertion failure.  A worker thread
%tries to acquire a mutex after it has been deleted by the main thread during
%shutdown, causing the assertion to fail.  
%
%Last writer slices provide the key piece of information to debug this failure.
%When the program fails, the core dump shows that the mutex is invalid because
%it has been deleted.  The mutex's last writer slice shows that the last write
%to the mutex before the worker tried to acquire it was when the main thread
%deleted it.  The last writer slice reveals the connection between the the two
%pieces of code involved in the bug: the code that acquires the mutex and the
%code that deletes the mutex.  Note that this connection is not obvious when the
%program crashes because the main thread may have executed arbitrarily far from
%where it deleted the mutex before the worker thread crashes.

@@

We used last writer slices to debug an atomicity violation in MySQL-4.0.12 that
is a critical security vulnerability.  Figure~\ref{fig:mysqllws} shows the
buggy code.  Thread 1 rotates the database's log file.  Log rotation should be
atomic, but the lack of synchronization allows Thread 2 to run its insert
during log rotation.  In that case, the insert query not logged, manifesting
the failure.  We added an {\tt abort} statement to the non-logging case because
this behavior is a security vulnerability that the bug report describes as
``critical''.  Adding the {\tt abort} is reasonable, as the bug report
describes the {\tt else} block as an error case.  Running in a debugger, a
programmer might equivalently use a breakpoint, rather than adding an
{\tt abort}.

@@

We triggered the bug by running an insert query during a log rotation and
execution stopped at {\bf C}.  The core dump showed the log was marked closed,
which is the failure state ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
However, the core dump did not reveal {\em why} the log was marked closed.  The
last writer slice at {\bf C} for {\tt log\_type} shows that it was written last
by Thread 1 at {\bf A}, in the log rotation code, which should be atomic.
Thus, the last writer slice shows the programmer the need for log rotation to
be atomic with respect to the insert code to prevent this failure.

@@

@@
